---
title: "Module 1: Analysis of the Haemopedia Database"
author: 
- name: Jason Cosgrove (jason.cosgrove@curie.fr)
  affiliation:  Institut Curie, UMR168
editor_options:
  chunk_output_type: inline
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---
---

<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>



# Introduction 
In this analysis we analyse the haemopedia bulk RNAseq database (1) to identify metabolic signature of lineage bias. More specifically, we perform data QC and normalisation followed by differential expression and pathway enrichment analysis.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
#load in the packages that we need to run this analysis
library(Seurat)
library(dplyr)
library(edgeR)
library(limma)
library(Glimma)
library(DESeq2)
library(cowplot)
library(ggplot2)
#package for volcano plot
library(ggrepel)
#package for pathway analysis
library(pathfindR)

source("helpermethods.R")
load("genesets/geneset_Robjects/metabolic_signatures.Rda")

#packages for pathfinder
BiocManager::install("KEGGREST")
BiocManager::install("KEGGgraph")
BiocManager::install("AnnotationDbi")
BiocManager::install("org.Hs.eg.db")


all.reg.immuno.genes <- read.csv("C:/Users/acer/Desktop/immuno_genes.csv", fileEncoding = "UTF-16") #regulatory immune genes dataset form GO term

all.immuno.genes <- read.csv("C:/Users/acer/Desktop/all_immuno_genes.csv", fileEncoding = "UTF-16") #all immune genes from GO terms

#set the seed so we can reproduce the results
set.seed(12345)


dir.create("Results",showWarnings = F)
dir.create("Results/Module1",showWarnings = F)

```


# Data QC and normalisation using the LIMMA package. 

Filter the dataset and normalise using limmas voom function. The voom method transforms count data to log2-counts per million (logCPM), then estimates the mean-variance relationship in the data to compute appropriate observation-level weights.


```{r LIMMA analysis, echo=FALSE, eval= TRUE}

#load in the haemopedia dataset which we have stored as a DGEobject, this has been downloaded. 
load("datasets/Haemopedia/DGE_object_haemopedia.Rda")


# filter out activated lymphocytes, megs and pDCs as these are not relevant to our research 
cells.to.remove <- c("GCB.1","GCB.2","SplPlsB.1" ,
                      "SplPlsC.1","SplPlsC.2",
                      "SplPlsC.3","BMPlsC.1","MemCD4T.1",
                      "EffCD4T.1","pDC.1",
                      "pDC.2",
                      "EffCD4T.2","MemCD4T.2",  
                      "MemCD8T.2","MemCD8T.1",
                      "MemCD4T.1", "pDC.1", "pDC.2",
                      "MegTPO.1",  "MegTPO.2")

#filter out lowly expressed genes, and also cells that we are not interested in. 
haemopedia.filtered <- filterData(haemopedia,cells.to.remove)

#add the lineage annotations for our analysis
lin <- performLineageAnnotation(haemopedia.filtered)


condition <-factor(lin)
design <- model.matrix( ~ 0 + condition)
colnames(design) <- levels( condition )


#calculate normalisation factors for each cell type
y <- calcNormFactors(haemopedia.filtered) 

#run the zoon transformation
v <- voom(haemopedia.filtered,design,plot = TRUE, span = 0.1,normalize.method = "quantile") 
#plot the library sizes for each sample as a sanity check
boxplot(v$E)
```



# Data visualisation using the Seurat package
Convert the data into a seurat object (3) and visualise using PCA and non-linear equivalents such as UMAP


```{r convertToSeurat, echo=FALSE, eval= TRUE, message = FALSE, warning = FALSE}

haem.seurat <- convertToSeurat(v$E)

#now lets run dimensionality reduction and visualisation to see if we have any batch effects in our data
haem.seurat <- RunPCA(object = haem.seurat, npcs = 50, verbose = FALSE)
haem.seurat  <- RunUMAP(object = haem.seurat, reduction = "pca", dims = 1:50, min.dist = 0.4, spread = 0.4,verbose = FALSE)


#visualise pca based dimensionality reduction vs. umap, this is done on all genes!
p1 <- DimPlot(object = haem.seurat, reduction = "pca", group.by = "lineage", pt.size = 1.6, label.size = 4,  label = F, repel = T) 
p2 <- DimPlot(object = haem.seurat, reduction = "umap", group.by = "lineage", pt.size = 1.6, label.size = 4,  label = F, repel = T)

p1
p2

plot_grid(p1, p2)


#save the seurat object to file for later analysis if we need it.
save(haem.seurat, file = "datasets/Haemopedia/haem_seurat.Rda")
```

# Mitochondrial genes
As we are interested in metabolism it is interesting to know how many of our genes
encode for proteins that localise to the mitochondria using the mitocarta v2 database (4). 


```{r mito, eval = FALSE}

mitocarta.genes <- read.csv("genesets/mitochondrial_assoc_genes.csv",header=F)

mito.genes <- intersect(rownames(haem.seurat),mitocarta.genes$V1)
haem.seurat <- AddModuleScore(haem.seurat, features = list(mito.genes),name = "mito") 
VlnPlot(haem.seurat, features = "mito1",group.by = "lineage")

mito <- haem.seurat@meta.data$mito1
names(mito) <- colnames(haem.seurat)

boxplot(mito[c("STHSC.1","STHSC.2")],mito[c("MPP.1","MPP.2")]  ,mito[c("LSK.1","LSK.2")]  ,  mito[c("CLP.1","CLP.2")] , mito[c("MEP.1","MEP.2")] , mito[c("GMP.1","GMP.2")])

```


# Differential expression analysis
Lets visualise our differentially expressed genes using a volcano plot. Overlay the expression of metabolic genes.

```{r pressure2}

#specify the contrast matrix that we want to use for our analysis
cont.matrix <- makeContrasts( E = Erythroid - Myeloid,
                              M = Myeloid - Erythroid,
                              levels = design)


#Now that we have the voom transformed data we can use limma to test for differential expression. First we fit a linear model for each gene using the lmFit function in limma. lmFit needs the voom object and the design matrix that we have already specified, which is stored within the voom object.
fit <- lmFit(v)

#Now we can apply the contrasts matrix to the fit object to get the statistics and estimated parameters of our comparison that we are interested in. 
fit.cont <- contrasts.fit(fit, cont.matrix)

#The final step is to call the eBayes function, which performs empirical Bayes shrinkage on the variances, and estimates moderated t-statistics and the associated p-values.
fit.cont <- eBayes(fit.cont, robust = T, proportion = 0.4)

#print a summary of the results
summa.fit <- decideTests(fit.cont)
summary(summa.fit)

#set the contrast names for when we plot the results
contrast.names <- c("E","M")

# get all metabolic genes
metabolic.genes <- c(metabolic.signatures$allGenesTested)

#get regulatory immune genes

reg.immuno.genes <- (all.reg.immuno.genes[,2])

reg.immuno <- intersect(rownames(haem.seurat),reg.immuno.genes)


#get all immuno genes

immuno.genes <- (all.immuno.genes[,2])

immuno <- intersect(rownames(haem.seurat), immuno.genes)

# do volcano plots for each lineage, overlaying the expression of all metabolic genes
res.E <- topTable(fit.cont, coef=1, number=nrow(v$E), adjust.method="BH")
res.E <- res.E[order(res.E$adj.P.Val),]
nsign.E <- length(which(res.E$adj.P.Val < 0.05))
nsignfc.E <- length(which(res.E$adj.P.Val < 0.05 & abs(res.E$logFC) > 1))
LimmaVolcano(res.E)


#create dataset for labeling immune regulation genes in volcano plot
test <- data.frame(genes = reg.immuno)

test.subset.E <- subset(res.E,genes %in% test$genes)

test.subset.M <- subset(res.M,genes %in% test$genes)

#volcano plot with labels
limma.volcano.immuno <- function(res, main="", fct=2, pt=0.05){
  res$sign <- 0
  res$sign[which(res$adj.P.Val < pt & abs(res$logFC) > fct)] <- 1
  
  res$sign[which(res$adj.P.Val < pt & abs(res$logFC) > fct & res$genes %in% reg.immuno)] <- 3
  
  res.fitered <- subset(res,genes %in% test$genes)
  
  p <- ggplot(data=res, aes(x=logFC, y=-log10(adj.P.Val), colour=as.factor(sign))) + geom_point( size=2) +
    
    scale_color_manual(name="", values=c("4" = "orange","3" = "blue" ,"2" = "red","1"=
                                           rgb(80/255,80/255, 80/255,0.2), "0"=rgb(220/255,220/255, 220/255,0.2))) +  
    ggtitle(paste0("Volcano Plot - Limma immuno", main)) +
    theme(legend.position = "none") + xlim(-12,12) + 
    xlab("log2 fold change") + ylab("-log10 adj pvalue") + 
    geom_vline(xintercept=c(-fct, fct), linetype=2) + 
    geom_hline(yintercept=-log10(pt), linetype=2)+
    geom_text_repel(data=filter(res.fitered, -log10(adj.P.Val)>20), aes(label=genes))
  p
}


limma.volcano.immuno(res.E)

res.E.filtered <- res.E[res.E$adj.P.Val < 0.05,]

res.M <- topTable(fit.cont, coef=2, number=nrow(v$E), adjust.method="BH")
res.M <- res.M[order(res.M$adj.P.Val),]
nsign.M <- length(which(res.M$adj.P.Val < 0.05))
nsignfc.M <- length(which(res.M$adj.P.Val < 0.05 & abs(res.M$logFC) > 1))
LimmaVolcano(res.M)

limma.volcano.immuno(res.M)

res.M.filtered <- res.M[res.M$adj.P.Val < 0.05,]

#heatmap for M filtered genes
res.M.filtered <- res.M[res.M$adj.P.Val < 0.05 & res.M$logFC > 0,]

genes <- c(res.M.filtered$gene[1:20])

coolmap(v[genes,])

#heatmap for E filtered genes
res.E.filtered <- res.E[res.E$adj.P.Val < 0.05 & res.E$logFC > 0,]

genes <- c(res.E.filtered$gene[1:20])

coolmap(v[genes,])



#heatmap of immune filtered genes in E?

test.subset.heatmap <- test.subset[test.subset$adj.P.Val < 0.05 & test.subset$logFC > 0,]

genes.E.immuno <- c(test.subset.heatmap$gene[1:30])

coolmap(v[genes.E.immuno,], cexRow=0.5)

#heatmap of immune filtered genes M

test.subset.M.heatmap <- test.subset.M[test.subset.M$adj.P.Val < 0.05 & test.subset.M$logFC > 0,]

genes.M.immuno <- c(test.subset.M.heatmap$gene[1:30])

coolmap(v[genes.M.immuno,], cexRow=0.5)


test.dataset.E.go <- data.frame(select(res.E, genes, logFC, adj.P.Val)) 

run_pathfindR(test.dataset.E.go, gene_sets = "GO-All", p_val_threshold = 0.05, adj_method = "bonferroni", output_dir = "C:/Users/acer/Desktop" )

write.csv(res.E.filtered, "Results/Module1/DEGs_E.csv")
write.csv(res.M.filtered, "Results/Module1/DEGs_M.csv")


 ```


# Gene-set overrepresentation analysis
Approach 1. Geneset overrepresentation analysis: This works by seeing if your differentially expressed genes occur at a high frequency in a given geneset, relative to chance. The Null hypothesis here is that if a geneset represents 1% of all genes in the experiment then 1% of the genes in the pathway would be differentially expressed. This does assume that genes are independent which has been shown not to be true.  We perform these analyses using the kegga and goana methods available in the limma package. 


```{r pressure4, eval = FALSE}

for(i in 1:length(contrast.names)){
  
  go <- goana(fit.cont, coef=i,species = "Mm",geneid = haemopedia$entrezgene[rownames(haemopedia$counts) %in% rownames(v$E)])
  kegg <- kegga(fit.cont, coef=i,species = "Mm",geneid = haemopedia$entrezgene[rownames(haemopedia$counts) %in% rownames(v$E)])
  
  go$FDR.up <- p.adjust(go$P.Up, method = "BH", n = length(go$P.Up))
  go$FDR.down <- p.adjust(go$P.Down, method = "BH", n = length(go$P.Down))
  
  kegg$FDR.up <-  p.adjust(kegg$P.Up, method = "BH", n = length(kegg$P.Up))
  kegg$FDR.down <- p.adjust(kegg$P.Down, method = "BH", n = length(kegg$P.Down))

  kegg <- kegg[kegg$FDR.up < 0.1 | kegg$FDR.down < 0.1,]
  go <- go[go$FDR.up < 0.1 | go$FDR.down < 0.1,]
  write.csv(go,   paste( "Results/Module1/" ,contrast.names[i],"GOANA_GO_pathways.csv",sep="_"))
  write.csv(kegg, paste( "Results/Module1/" ,contrast.names[i],"KEGGA_KEGG_pathways.csv",sep="_"))
}



go.table <- read.csv("Results/Module1/_E_GOANA_GO_pathways.csv")

go.table.pval.ordered <- go.table[with(go.table, order("P.Up")), ]

go.table.pval.ordered <-go.table %>% arrange(desc(-FDR.up))

barplot(go.table.pval.ordered$FDR.up[1:10])

```


Approach 2. Gene set testing: We run two different types of tests: Competitive and self contained, that differ to the genset overrepresentation analysis. Competitive gene set tests, such as those implemented in GOseq and camera (5), ask the question whether the differentially expressed genes tend to be over-represented inthe gene set, compared to all the other genes in the experiment.Self-contained tests, which include the ROAST procedure, ask the question are the genes in the set/pathway differentially expressed as a whole. ROAST tends to be used for smaller amounts of pathways ROAST (6) (FRY is a fast approximation) uses different metrics to summarise geneset expression, we use the default of mean, this option is not available in camera. It works under three different hypotheses, "Up" to test whether the genes in the set tend to be up-regulated, with positive t-statistics, "Down" to test whether the genes in the set tend to be down-regulated, with negative t-statistics, and "Mixed" to test whether the genes in the set tend to be differentially expressed, without regard for direction.
Competitive null: The genes in the gene set do not have stronger association with the subject condition than other genes. (Approach 1, 2)
Self-contained null: The genes in the gene set do not have any association with the subject condition (Approach 3).

```{r pressure3, eval = FALSE}

#get the indices of the dataset that correspond to each geneset
load("genesets/geneset_Robjects/KEGG.Rda")
KEGG.ind <- ids2indices(KEGG.sets, rownames(v))

for(i in 1:length(contrast.names)){
  
  print(contrast.names[i])
  
  KEGG.camera <- camera(v,index = KEGG.ind,design=design,contrast = cont.matrix[,i],inter.gene.cor=0.01)
  KEGG.fry <- mroast(v, index = KEGG.ind, design = design, contrast = cont.matrix[,i], sort = "mixed",set.statistic = "mean50")
  
  KEGG.camera <- KEGG.camera[KEGG.camera$FDR < 0.1,]
  KEGG.fry <- KEGG.fry[KEGG.fry$FDR < 0.1,]
  
  write.csv(KEGG.camera, paste( "Results/Module1/" , contrast.names[i],"CAMERA_KEGG_pathways.csv",sep="_"))
  write.csv(KEGG.fry, paste( "Results/Module1/"  , contrast.names[i],"FRY_KEGG_pathways.csv",sep="_"))
}


```



# References: 
1. Choi, J. et al. “Haemopedia RNA-Seq: A Database of Gene Expression during Haematopoiesis in Mice and Humans.” Nucleic Acids Research 47, no. D1 (January 8, 2019): D780–85. https://doi.org/10.1093/nar/gky1020.

2. Ritchie, Matthew E. et al. “Limma Powers Differential Expression Analyses for RNA-Sequencing and Microarray Studies.” Nucleic Acids Research 43, no. 7 (April 20, 2015): e47–e47. https://doi.org/10.1093/nar/gkv007.

3. Butler, Andrew et al. Integrating Single-Cell Transcriptomic Data across Different Conditions, Technologies, and Species. Nature Biotechnology 36, no. 5 (May 2018). https://doi.org/10.1038/nbt.4096.

4. Calvo, Sarah E., Karl R. Clauser, and Vamsi K. Mootha. “MitoCarta2.0: An Updated Inventory of Mammalian Mitochondrial Proteins.” Nucleic Acids Research 44, no. D1 (January 4, 2016): D1251–57. https://doi.org/10.1093/nar/gkv1003.

5. Wu, Di, and Gordon K. Smyth. “Camera: A Competitive Gene Set Test Accounting for Inter-Gene Correlation.” Nucleic Acids Research 40, no. 17 (September 1, 2012): e133–e133. https://doi.org/10.1093/nar/gks461.

6. Wu, Di, et al. “ROAST: Rotation Gene Set Tests for Complex Microarray Experiments.” Bioinformatics 26, no. 17 (September 1, 2010): 2176–82. https://doi.org/10.1093/bioinformatics/btq401.
