---
title: "Module 3: Machine learning and Network Analysis of Metabolic Identities"
author: 
- name: Jason Cosgrove (jason.cosgrove@curie.fr), Vincent Cabeli
  affiliation:  Institut Curie, UMR168
editor_options:
  chunk_output_type: inline
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---
---



<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>



# Introduction 
In this analysis we develop a supervised learning classifier to predict lineage from metabolic state, and find out which genes are the most influential in this classification task. Lastly, we model gene-gene interactions using a causal network approach to understand how genes within our lineage specific metabolic signatures relate to each other.



```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#clear the workspace and set the working directory
rm(list = ls())

#load in the packages that we need to run this analysis
library(Seurat)
library(xgboost)
library(pROC)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
source("helpermethods.R")


dir.create("Results",showWarnings = F)
dir.create("Results/Module3",showWarnings = F)

#set the seed so we can reproduce the results
set.seed(12345)

```



load in gene sets that are relevant to our research question
```{r genesets}

load("genesets/geneset_Robjects/metabolic_signatures.Rda")


```



# Classifier training

In this section we train a classifier to predict lineage from gene expression profiles. Specifically, we train an xgboost model on just mature cells
using k-folds cross validation. 

```{r ClassifierTraining}


#load in teh haemopedia dataset in the form of a seurat object
load("datasets/Haemopedia/haem_seurat.Rda")


lskVariable100100 <- replicate(100, array(sample(intersect(metabolic.signatures$lskVariable,metabolic.signatures$allGenesTested), 100, replace = F )))

testingallgenes <- replicate(100, sample(rownames(haem.seurat), 100, replace = F ))

AreaucDataTest100 <- array()

AreaucDataTestallgenes <- array()


for (i in 1:ncol(testingallgenes)) {

  interesting.genes <- testingallgenes[,i]

  interesting.genes <- intersect(interesting.genes, rownames(haem.seurat))
  
  print(interesting.genes)
  
#prep the inputs for xgboost, im tempted to retry with the progenitors included in the training just to see how that translates to the single cell
  mpponly <- FALSE #do you want to include RPPs in the training? if so set to true 
  data.withlineage <- prepData(interesting.genes,haem.seurat, mpponly)

  data.withlineage$lin <- droplevels(data.withlineage$lin)
  labels = data.withlineage$lin
  data = as.matrix(data.withlineage[,names(data.withlineage) != 'lin'])
  
  
  # convert the data matrix to an xgb.matrix object
  xgb_data = xgb.DMatrix(data, label=as.numeric(labels)-1)
  
  
  
  weights <- sapply(labels, calculateLineageWeights)
  
  
  
  # Train model with cross validation
  # See https://xgboost.readthedocs.io/en/latest/parameter.html for full list of parameters
  
  nfolds = 10   # CV folds, should be at least 10, that gives 8 cells or so per fold and makes sure that underrepresented classes (erythroid; 15 cells) are not completely in the testing folds
  nrounds = 12  # boosting iterations; was 20
  
  set.seed(178352323) #you must set the seed here, if you set in the beginning but the ordering of operations changes, then the randonm number generator will not be equivalent!
  
  cv = xgb.cv(data = data, label = as.numeric(labels)-1, nfold = nfolds,
              params = list(subsample=0.3, colsample_bytree=0.3,colsample_bynode = 0.3,weight = weights),
              nrounds = nrounds,
              prediction = T, verbose = T,
              metrics = 'merror', objective = "multi:softprob", num_class = length(unique(labels)),
              callbacks = list(cb.cv.predict(save_models = TRUE)))
  xgb.preds = apply(cv$pred, 1, which.max) 
  
  predictiontable <- table(xgb.preds, labels)
  predictiontable
  
  roc_obj<- multiclass.roc(labels, xgb.preds)
  
  areauc <- auc(roc_obj)
  
  performance <- assessPerformanceOnPrecursors(haem.seurat,cv)
  performance
  
  AreaucDataTestallgenes <- append(AreaucDataTestallgenes, areauc)
  
  }
  
AreaucDataTestallgenes#results from 100 sets by 100 gens in all genes in haem.seurat dataset
  
AreaucDataTest100#resuts from 100 sets by 100 genes in intersect of lsk and metabolic genes


table(AreaucDataTestallgenes)[names(table(AreaucDataTestallgenes)) == 1]#count the number of auc == 1, ==6

table(AreaucDataTest100)[names(table(AreaucDataTest100)) ==1] # count the number of auc == 1, ==35
  
write.csv(AreaucDataTestallgenes, file = "Results/Module3/classifierModelAreaucDataTestallgenes.csv", append = T, row.names = T, col.names = T, sep = ",")

write.csv(AreaucDataTest100, file = "Results/Module3/classifierModelAreaucDataTest100.csv", append = T, row.names = T, col.names = T, sep = ",")

library(ggpubr)



AllGenesplot <- data.frame(auc = AreaucDataTestallgenes[2:101], geneset='Allgenes')

LSKplot <- data.frame(auc = AreaucDataTest100[2:101], geneset='LSKgenes')

plotdataAucTest <- rbind(AllGenesplot, LSKplot )

aucboxplot <- ggplot(plotdataAucTest, aes(x = geneset, y = auc)) + geom_boxplot() 

aucboxplot +stat_compare_means()

png(filename="your/file/location/name.png")
plot(fit)
dev.off()

  
```
  
  
  # Classifier Evaluation
  We then see how our classifier performs when we get it to predict the lineage of early lineage-biased bone marrow precursors.
  
  ```{r ClassifierEvaluation}
  # xgboost actually otputs a dataframe where each column represents a lineage each row is an individual sample and each entry is the prob that the sample belongs to a given lineage
  xgb.preds = apply(cv$pred, 1, which.max) 
  
  predictiontable <- table(xgb.preds, labels)
  predictiontable
  
  roc_obj<- multiclass.roc(labels, xgb.preds)
  
  areauc <- auc(roc_obj)
  numericAuc <- as.numeric(areauc)

#see how well we can predict the fate of precurosrs

  performance <- assessPerformanceOnPrecursors(haem.seurat,cv)
  performance
  
  AreaucDataTest <- append(AreaucDataTest, areauc)
  PerfonamceDataTest <- append(PerfonamceDataTes, performance)



AreaucDataTest

PerfonamceDataTest

write.csv(TableDataTest, file = "Results/Module3/classifierModeltestareauc.csv", append = T, row.names = T, col.names = T, sep = ",")



save(cv,file = "Results/Module3/classifierModel.Rda")

```

# Calculate SHAPLEY Scores
Using shapley scores quantify which genes are the most important in informing our classifiers lineage predictions. 

```{r calculateSHAPLEYScores, eval = TRUE}


# when you call xgbimportance on the model you get all genes used by the model
all_features = lapply(cv$models, function(x) {xgb.importance(model = x, feature_names = colnames(data))}$Feature)
all_features = unique(unlist(all_features))

#create a dictionary that links lineage to its numeric entry in our classifier model
labels.ordered <- colnames(table(xgb.preds, labels))
dict <- vector(mode="list", length=length(labels.ordered))
names(dict) <- 1:(length(labels.ordered))
for(i in 1:length(labels.ordered)){dict[[i]] <- labels.ordered[i]}
  
SHAP_results <- calculateSHAPScores(all_features,dict,3,data)

shapscores.summarised <- createSHAPSummary(SHAP_results)



seq <- c(metabolic.signatures$sequential_erythroid,
         metabolic.signatures$sequential_myeloid,
         metabolic.signatures$sequential_lymphoid)

#write the data out into a format that is compatible with miic
write.csv(data.withlineage[,c(rownames(shapscores.summarised),"lin")],"Results/module3/machine_for_miic.csv")
write.csv(data.withlineage[,c(seq,"lin")],"Results/module2/sequential_for_miic.csv")

```


# Summarise SHAPLEY Scores
Using shapley scores quantify which genes are the most important in informing our classifiers lineage predictions. 

```{r summariseSHAPLEYScores, eval = TRUE}


shapscores.summarised <- data.frame(shapscores.summarised)
shapscores.summarised.ordered <- shapscores.summarised[order(shapscores.summarised$SHAP_E,
                                    shapscores.summarised$SHAP_M,
                                    shapscores.summarised$SHAP_L,decreasing = T),]

pheatmap(abs(shapscores.summarised.ordered),cluster_cols = FALSE,cluster_rows = F,
         fontsize_row = 5,
         colorRampPalette(rev(brewer.pal(n = 11, name ="RdYlBu")))(100))

```

# Cluster Informative Genes
Using shapley scores quantify which genes are the most important in informing our classifiers lineage predictions. 


```{r clusterSigs, eval = TRUE}
#based on these genes we need to do kmeans clustering, we know how many liineages so we can set k. get the kmeans code from the rnaseq tutorial



x <- AverageExpression(haem.seurat, features = rownames(shapscores.summarised),add.ident = "lineage")

x <- x$RNA[,colnames(x$RNA) %in% c("haemopedia_Myeloid","haemopedia_Erythroid","haemopedia_Lymphoid")]

clusters <- colnames(x)[apply(x,1,which.max)]
names(clusters) <- rownames(shapscores.summarised)

e.sig <- names(clusters[clusters == "haemopedia_Erythroid"])
m.sig <- names(clusters[clusters == "haemopedia_Myeloid"])
l.sig <- names(clusters[clusters == "haemopedia_Lymphoid"])



#order the genes by the kmeans as this will be useful for visualisaßtion

genes <- c(e.sig, m.sig,l.sig)

#in the final pipeline we want this figure included
cells <- colnames(haem.seurat)[haem.seurat@meta.data$lineage %ni% c("MPP", "Meg","RPP")]

genes <- c("Abcb10","Tbxas1","Gsr",
           "Atp6ap2","Idh1","Hk3","Tmem14c",
           "Uqcr11","G6pdx","Hexa","Gpx1",
           "Gstm5","Pkm","Atpif1","Slc14a1",
           "Hk2","Neu1","Coxb2","Mgat4b",
           "Pol2ra")

x <- DoHeatmap(haem.seurat,features = genes, group.by = "lineage", cells = cells,disp.min = -1.5, disp.max = 1.5)

x + theme(axis.text.y = element_text(size=5)) + scale_fill_gradient2(low = "white",mid = "light grey",
     high = "dark blue")

```

# Generate Lineage Signatures
Using shapley scores quantify which genes are the most important in informing our classifiers lineage predictions. 

```{r featureSelection, eval = TRUE}
dataset <- as.matrix(haem.seurat@assays$RNA@counts)

#discard genes which dont follow the trend at the level of RPPs
m.sig.filter2 <- filterOnBulkProgenitors(m.sig,c("GMP.1","GMP.2"),c("CLP.1","CLP.2","MEP.1","MEP.2"),dataset)
e.sig.filter2 <- filterOnBulkProgenitors(e.sig,c("MEP.1","MEP.2"),c("CLP.1","CLP.2","GMP.1","GMP.2"),dataset)
l.sig.filter2 <- filterOnBulkProgenitors(l.sig,c("CLP.1","CLP.2"),c("GMP.1","GMP.2","MEP.1","MEP.2"),dataset)


metabolic.signatures$erythroid_machinelearning <- e.sig.filter2
metabolic.signatures$myeloid_machinelearning <- m.sig.filter2
metabolic.signatures$lymphoid_machinelearning <- l.sig.filter2


save(metabolic.signatures,file = "genesets/geneset_Robjects/metabolic_signatures.Rda")


```

